Computational biology in the 21st century:\\Scaling with compressive algorithms

### 2015-09-20 main.tex
Very compressed single network science section

When examining more complex phenomena than single proteins or compounds, we
often look to synthesize things together.
To that end, we frequently use
networks to represent biological data, such as the genetic
and physical interactions among proteins, as well as those in metabolic 
pathways~\cite{berger2013computational}.
These relationships and interactions play crucial roles in a systems-level understanding of biology.
While standard network science tools have been employed in these analyses, metabolic pathways have additional flow constraints on their links;
as with our other examples of interesting biological data types, these additional constraints on metabolic pathways allows for more targeted approaches to gain advantages over more general techniques.

\section{Challenges with biological data}

@@

With biological networks, challenges lie in integrating such data in order to 
deduce gene function, as well as uncover disease and metabolic pathways.
Several approaches have taken advantage of the particular topologies of 
biological
networks, such as diffusion-based approaches, which explore the topology of
networks through random walks.
In protein interaction networks, some nodes are high-degree \emph{hubs,} which 
interact with
many partners; these interactions do not imply shared function as strongly as
non-hub interactions; such information needs to be teased apart.
As new functional annotations and interactions are discovered,
the number and size of these networks grows, making related problems more 
intractable~\cite{berger2013computational}.

@@

An area of great interest in integrating biological network data is in global multiple network 
alignment, which is based on network topology and conserved biological function (IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn}).
IsoRank's approach is similar to the PageRank algorithm~\cite{berger2013computational} 
but performs a random walk on the product graph of two networks;
IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
PageRank-Nibble algorithm~\cite{berger2013computational}.
NetworkBLAST~\cite{atias2012comparative} is another approach which searches for 
conserved paths and conserved interaction clusters.
Diffusion state distance (DSD)~\cite{cao2013going} is a diffusion-based approach
which defines a distance metric between two nodes based on random walks.
DSD penalizes high-degree nodes, and outperforms other graph-theoretic distance
functions when applied to standard approaches such as neighbor voting for function annotation.
Diffusion component analysis (DCA)~\cite{cho2015diffusion} amounts to a network-based 
analogue of principle component analysis; DCA reduces the dimensionality of a
network in such a way that function annotation can be performed in the reduced 
space, significantly decreasing running time while still providing competitive
prediction accuracy.
Using a compressed representation of metabolic networks, 
Mongoose~\cite{chindelevitch2014exact} is able to 
identify and unblock many blocked reactions--those whose metabolites are out of 
balance---in current models without loss of accuracy.

>>>>

When examining more complex phenomena than single proteins or compounds, we
often look to synthesize things together into a systems-level understanding of biology.
To that end, we frequently use
networks to represent biological data, such as the genetic
and physical interactions among proteins, as well as those in metabolic 
pathways~\cite{berger2013computational}.
While standard network science tools have been employed in these analyses---
e.g. several approaches make use of diffusion/random walks to explore the topology of networks \cite{cao2013going,cho2015diffusion}---
they are often paired with more specific biological data,
as seen in IsoRank~\cite{singh2008global} and IsoRankN's~\cite{liao2009isorankn} use of conserved biological function in addition to random walks for global multiple network alignment.
Other tools solve more specifically biological problems, such as Mongoose~\cite{chindelevitch2014exact}, which 
aims to identify and unblock metabolic networks.
However, given its breadth, biological network science is beyond the scope of this Review, so we will not focus on that here.

%NetworkBLAST~\cite{atias2012comparative} is another approach which searches for 
%conserved paths and conserved interaction clusters.
%IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
%PageRank-Nibble algorithm~\cite{berger2013computational}.
%Diffusion state distance (DSD)~\cite{cao2013going} is a diffusion-based approach
%which defines a distance metric between two nodes based on random walks.
%DSD penalizes high-degree nodes, and outperforms other graph-theoretic distance
%functions when applied to standard approaches such as neighbor voting for function annotation.
%Diffusion component analysis (DCA)~\cite{cho2015diffusion} amounts to a network-based 
%analogue of principle component analysis; DCA reduces the dimensionality of a
%network in such a way that function annotation can be performed in the reduced 
%space, significantly decreasing running time while still providing competitive
%prediction accuracy.
\section{Challenges with biological data}

@@

@@
